Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-03-03 ThesisJP/Background/background.tex
Small changes in thesis background.

For the first adopters general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power. Balancing typically means that highly parallel and computationally intense tasks should be offloaded to the GPU and serialized and memory intensive work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics. Even if an on-chip GPU is less powerful than a high-performance discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard. This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web-service and automatically scales on demand. Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}). In this thesis we focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is design to be the main control unit of a computer. It is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see what these similarities are and why there is still a big difference in peak performance between a CPU and a GPU.

@@

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}). This extension has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instruction set have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512). AVX and AVX2 adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports supporting FMA-AVX2 instructions, and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture. In addition, the most powerful integrated GPU found on Haswell CPUs (HD 5200) have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz. This adds
\begin{equation}
16*40*1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors (Fig.\,\ref{fig:cpu_gpu}). The Kepler architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The theoretical peak single precision FLOPS for this architecture is by that found to be 
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of five TFLOPS in computationally power as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with SIMD cores on the CPU, results in five times the computational performance of a high-end CPU (without the integrated GPU).

@@

Both CUDA and OpenCL is centered around a kernel function which are executed by a large group of threads. The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers. A CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}. The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA, this kernel function is launched by using a CUDA specific syntax:
\begin{align*}
\texttt{square<<<grid, block>>>(a, N)},
\end{align*}
where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the buffer size. When launched, a copy of the kernel is scheduled  for each thread in each block, and each kernel get as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}). This information is then typically used to calculate which memory elements to gather from the global memory buffer \texttt{a}. In order to formulate a given problem as a kernel function the global memory input to each thread must not rely on the output from other threads, hence it must be possible for the GPU to scheduled threads in any order.

@@

No doubt, it is still a major overhead involved with writing GPU code, and in the future we will probably see work aimed at reducing this overhead.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in the future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions is handled in CPU code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the problem at hand are often able to make faster code by manually writing SIMD instruction code. The relevance of auto-applied GPU instructions by compilers will also increasing due to the recent embedding of GPUs on the same chip as the CPU. Soon, when writing code, we will probably talk about coding for a collection of heterogeneous cores where operations are automatically offloaded to special cores (e.g. a GPU) by the compiler.

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used, where frequency dependent attenuation force us to select a lower frequency if the object we want to image is located deep into the body. This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging. A piezoelectric ceramic is used to generate the ultrasonic frequencies, and is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically design for each medical imaging modality. For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs. A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo. This amplitude trace is known as an A-mode image.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse integrated in time.} of ultrasound by diffraction. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}
\delta\theta = \frac{\lambda}{D},
\end{align}
\nomi{$\delta\theta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture width} is the aperture width. With a speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution of a 2 cm cardiac array at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space. This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased arrays \cite{AngelUltrasound}. A cardiac ultrasound array is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$, resulting in sector which are $\theta$ wide. This kind of image is known as a B-mode image (B for brightness). If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array weight vector or apodization, and $z[n]$ is the output A-mode image.  If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_m+p| - |p_o+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays have to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays are therefore not time varying.

>>>>

For the first adopters of general-purpose computing on GPUs (GPGPU) it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power. Balancing typically means that highly parallel and computationally intense tasks should be offloaded to the GPU and serialized and memory intensive work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics. Even if an on-chip GPU is less powerful than a high-performance discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard. This is the reason why one of the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU memory transfers.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web service and automatically scales on demand. Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}). In this thesis we focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is design to be the main control unit of a computer, it is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see what these similarities are and why there is still a big difference in peak performance between a CPU and a GPU.

@@

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}). This extension has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instruction set have added new instructions and longer registers (SSE2, SSE3, SSE4, \nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512). The newly introduced extensions, AVX and AVX2, adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports supporting FMA-AVX2 instructions, and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture. In addition, the most powerful integrated GPU found on Haswell CPUs, the HD 5200, have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz. This adds
\begin{equation}
16*40*1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors (Fig.\,\ref{fig:cpu_gpu}). The Kepler architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The theoretical peak single precision FLOPS for this architecture is by that found to be 
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of five TFLOPS as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with SIMD cores on the CPU, results in five times the computational performance of a high-end CPU (without the integrated GPU).

@@

Both CUDA and OpenCL is centered around a kernel function which are executed by a large group of threads. The kernel function, for both CUDA and OpenCL, has a C-like syntax with some additional specifiers. A CUDA kernel is depicted in Fig.\,\ref{fig:gpu_kernel}. The \texttt{\_\_global\_\_} specifier tells the CUDA compiler that this function is a kernel function, and in CUDA, this kernel function is launched by using a CUDA specific syntax:
\begin{align*}
\texttt{square<<<grid, block>>>(a, N)},
\end{align*}
where \texttt{grid} and \texttt{block} is 3-component vectors specifying grid and block size in "xyz" as depicted in Fig.\,\ref{fig:gpu_grid}, \texttt{a} is a pointer to a memory buffer located on the GPU, and \texttt{N}$*$\texttt{N} is the buffer size. When launched, a copy of the kernel is scheduled  for each thread in each block, and each kernel gets as input its location in the compute grid (\texttt{blockIdx} and \texttt{threadIdx}) together with the grid size (\texttt{gridDim} and \texttt{blockDim}). This information is then typically used to calculate which memory elements to gather from global memory, in this case the elements in \texttt{a}. In order to formulate a given problem as a kernel function the global memory input to each thread must not rely on the output from other threads, hence it must be possible for the GPU to scheduled threads in any order.

@@

No doubt, it is still a major overhead involved with writing GPU code, and in the future we will probably see work aimed at reducing this overhead.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in the future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions is handled in CPU code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the problem at hand are often able to make faster code by manually writing SIMD instruction code. The relevance of auto-applied GPU instructions by compilers will also increase due to the recent embedding of GPUs on the same chip as the CPU. Soon, when writing code, we will probably talk about coding for a collection of heterogeneous cores where operations are automatically offloaded to special cores (e.g. a GPU) by the compiler.

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used, where frequency dependent attenuation in human tissue force us to select a lower frequency if the object we want to image is located deep into the body. This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging. A piezoelectric ceramic is used to generate the ultrasonic frequencies, and is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically designed for each medical imaging modality. For cardiac ultrasound imaging the probe must not only be designed for a certain frequency range, its footprint also needs to be smaller than the intercostal window in order to avoid shadowing by ribs. A line in an ultrasound image is generated by firing a ultrasound pulse into the body and sample the received echo. This amplitude trace is known as an A-mode image.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted pulse will get shaped into a beam\footnote{A beam refers to the pulse integrated in time.} of ultrasound by diffraction. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}
\delta\theta = \frac{\lambda}{D},
\end{align}
\nomi{$\delta\theta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength, and \nom{$D$}{Aperture width} is the aperture width. With a speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution of a 2 cm cardiac array at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of around 2 mm. The axial resolution is inverse proportional to the pulse length, and is typically better than the lateral resolution.

If time delays are applied on the signal sent to the different elements it is possible to steer and focus the ultrasound beam in space. This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied both when transmitting and receiving ultrasound pulses. For medical ultrasound imaging there are two main types of ultrasound probes, linear and phased arrays \cite{AngelUltrasound}. A cardiac ultrasound array is a phased array, which means that the image is constructed from beams, or A-mode images, which are gradually steered from $-\theta/2$ to $\theta/2$, resulting in an imaging sector which are $\theta$ wide. This kind of image is known as a B-mode image (B for brightness). If \nom{$M$}{Number of array elements} is the number of elements in the array, and \nom{$\Delta_m[n]$}{Per-element focusing and steering delays} is the per-element focusing and steering delay, the standard form of beamforming used on receive, known as \textit{delay-and-sum}, is given as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where \nom{$\vec{w}$}{Array weight vector (apodization)} is know as the array weight vector or apodization, and $z[n]$ is the output A-mode image.  If $p_o$ is the phase center of our array and $p$ is the point where we want to steer and focus the ultrasound beam, we have to use the following delays:
\begin{align}
\Delta_m[n] =\frac{|p_m+p| - |p_o+p|}{c},
\end{align}
where \nom{$c$}{Speed of sound} is the speed of sound in human tissue, and $p_m$ is the position of the m'th element. On transmit the inverse set of delays have to be used. Also note that on transmit we have to decide where to steer and focus prior to transmitting a pulse, transmit delays are therefore not time varying for an A-mode image.
